---
title: "Lab 05 - Data Wrangling"
output: html_document
date: 2023-09-29
---

# Learning goals

-   Use the `merge()` function to join two datasets.
-   Deal with missings and impute data.
-   Identify relevant observations using `quantile()`.
-   Practice your GitHub skills.

# Lab description

For this lab we will be, again, dealing with the meteorological dataset downloaded from the NOAA, the `met`. In this case, we will use `data.table` to answer some questions regarding the `met` dataset, while at the same time practice your Git+GitHub skills for this project.

This markdown document should be rendered using `github_document` document.

# Part 1: Setup the Git project and the GitHub repository

1.  Go to your documents (or wherever you are planning to store the data) in your computer, and create a folder for this project, for example, "PM566-labs"

2.  In that folder, save [this template](https://raw.githubusercontent.com/USCbiostats/PM566/master/website/content/assignment/05-lab.Rmd) as "README.Rmd". This will be the markdown file where all the magic will happen.

3.  Go to your GitHub account and create a new repository, hopefully of the same name that this folder has, i.e., "PM566-labs".

4.  Initialize the Git project, add the "README.Rmd" file, and make your first commit.

5.  Add the repo you just created on GitHub.com to the list of remotes, and push your commit to origin while setting the upstream.

Most of the steps can be done using command line:

``` sh
# Step 1
cd ~/Documents
mkdir PM566-labs
cd PM566-labs

# Step 2
wget https://raw.githubusercontent.com/USCbiostats/PM566/master/website/content/assignment/05-lab.Rmd 
mv 05-lab.Rmd README.md

# Step 3
# Happens on github

# Step 4
git init
git add README.Rmd
git commit -m "First commit"

# Step 5
git remote add origin git@github.com:[username]/PM566-labs
git push -u origin master
```

You can also complete the steps in R (replace with your paths/username when needed)

``` r
# Step 1
setwd("~/Documents")
dir.create("PM566-labs")
setwd("PM566-labs")

# Step 2
download.file(
  "https://raw.githubusercontent.com/USCbiostats/PM566/master/website/content/assignment/05-lab.Rmd",
  destfile = "README.Rmd"
  )

# Step 3: Happens on Github

# Step 4
system("git init && git add README.Rmd")
system('git commit -m "First commit"')

# Step 5
system("git remote add origin git@github.com:[username]/PM566-labs")
system("git push -u origin master")
```

Once you are done setting up the project, you can now start working with the MET data.

## Setup in R

1.  Load the `data.table` (and the `dtplyr` and `dplyr` packages if you plan to work with those).

```{r}
library(dplyr)
library(data.table)
library(dtplyr)
library(ggplot2)
library(tidyverse)
library(leaflet)

```

2.  Load the met data from <https://raw.githubusercontent.com/USCbiostats/data-science-data/master/02_met/met_all.gz>, and also the station data. For the later, you can use the code we used during lecture to pre-process the stations data:

```{r stations-data, eval = TRUE}
# Download the data
stations <- fread("ftp://ftp.ncdc.noaa.gov/pub/data/noaa/isd-history.csv")
stations
stations[, USAF := as.integer(USAF)]

# Dealing with NAs and 999999
stations[, USAF   := fifelse(USAF == 999999, NA_integer_, USAF)]
stations[, CTRY   := fifelse(CTRY == "", NA_character_, CTRY)]
stations[, STATE  := fifelse(STATE == "", NA_character_, STATE)]

# Selecting the three relevant columns, and keeping unique records
stations <- unique(stations[, list(USAF, CTRY, STATE)])

# Dropping NAs
stations <- stations[!is.na(USAF)]

# Removing duplicates
stations[, n := 1:.N, by = .(USAF)]
stations <- stations[n == 1,][, n := NULL]

```

3.  Merge the data as we did during the lecture.

```{r}
# Where are we getting the data from
met_url <- "https://github.com/USCbiostats/data-science-data/raw/master/02_met/met_all.gz"
# Downloading the data to a tempfile (so it is destroyed afterwards)
# you can replace this with, for example, your own data:
# tmp <- tempfile(fileext = ".gz")
tmp <- "met.gz"
# We sould be downloading this, ONLY IF this was not downloaded already.
# otherwise is just a waste of time.
if (!file.exists(tmp)) {
  download.file(
    url      = met_url,
    destfile = tmp,
    # method   = "libcurl", timeout = 1000 (you may need this option)
  )
}

dat <- fread(tmp)
head(dat)

#Merging the Data Set 
merge(
  # Data
  x     = dat,      
  y     = stations, 
  # List of variables to match
  by.x  = "USAFID",
  by.y  = "USAF", 
  # Which obs to keep?
  all.x = TRUE,      
  all.y = FALSE
  ) %>% nrow()

stations[, n := 1:.N, by = .(USAF)]
stations <- stations[n == 1,][, n := NULL]

dat <- merge(
  # Data
  x     = dat,      
  y     = stations, 
  # List of variables to match
  by.x  = "USAFID",
  by.y  = "USAF", 
  # Which obs to keep?
  all.x = TRUE,      
  all.y = FALSE
  )
head(dat[, list(USAFID, WBAN, STATE)], n = 4)





```

## Question 1: Representative station for the US

What is the median station in terms of temperature, wind speed, and atmospheric pressure? Look for the three weather stations that best represent continental US using the `quantile()` function. Do these three coincide?

```{r}

 mean_of_data<- dat %>%
  group_by(USAFID) %>%
  summarize(    
    Mean_Temp = mean(temp, na.rm = TRUE),
    Mean_Wind = mean(wind.sp, na.rm = TRUE),
    Mean_ATM = mean(atm.press, na.rm = TRUE))
mean_of_data
  
m1<-quantile(mean_of_data$Mean_Temp, 0.5, na.rm=TRUE, type=1)
m2<-quantile(mean_of_data$Mean_Wind, 0.5, na.rm=TRUE, type=1)
m3<-quantile(mean_of_data$Mean_ATM, 0.5, na.rm=TRUE, type=1)

mean_of_data %>% filter(Mean_Temp ==m1 )
mean_of_data %>% filter(Mean_Wind ==m2 )
mean_of_data %>% filter(Mean_ATM ==m3 )
```

Knit the document, commit your changes, and push it to GitHub. Don't forget to add `README.md` to the tree, the first time you render it.

## Question 2: Representative station per state

Just like the previous question, you are asked to identify what is the most representative, the median, station per state. This time, instead of looking at one variable at a time, look at the euclidean distance. If multiple stations show in the median, select the one located at the lowest latitude.

```{r}
#d = √[ (x2 – x1 )2 + (y2 – y1 )2] 
representative_stations <- dat %>% group_by(STATE) %>%
  mutate(euclidean = sqrt((lat - mean(lat))^2 + (lon - mean(lon))^2)) %>% arrange(STATE, euclidean, lat) %>% slice(1) %>% select(STATE, USAFID, lat, lon, euclidean)
representative_stations


station_order <- representative_stations[order(representative_stations$lat, decreasing = TRUE),]
lowest_lat <- station_order[1, ]
lowest_lat
```

Knit the doc and save it on GitHub.

## Question 3: In the middle?

For each state, identify what is the station that is closest to the mid-point of the state. Combining these with the stations you identified in the previous question, use `leaflet()` to visualize all \~100 points in the same figure, applying different colors for those identified in this question.

```{r}
# Midpoint coordinates for each state: Group by state and find the mean of lat and lon
midpoints <- dat %>% group_by(STATE) %>% summarize(
    mid_lat = mean(lat),
    mid_lon = mean(lon))
midpoints

# Closest station to a given point
find_closest_station <- function(point, stations) {
  distances <- sqrt((stations$lat - point$mid_lat)^2 + (stations$lon - point$mid_lon)^2)
  closest_station <- stations[min(distances), ]
  return(closest_station)}

# Find the closest station to each midpoint
closest_stations <- midpoints %>% rowwise() %>% do(find_closest_station(., dat))

# Combine the closest stations with the most representative stations
all_stations <- rbind(representative_stations, closest_stations)

# Leaflet map  visualizes all stations ( couldn't figure out the color D:)
leaflet(data = all_stations) %>%
  addTiles() %>%
  addCircleMarkers(
    lng = ~lon,
    lat = ~lat,
    color = ~factor(STATE, levels = unique(all_stations$STATE)),
    group = "Stations",
    fillColor = ,
    fillOpacity = 0.7,
    radius = 3) 
```

Knit the doc and save it on GitHub.

## Question 4: Means of means

Using the `quantile()` function, generate a summary table that shows the number of states included, average temperature, wind-speed, and atmospheric pressure by the variable "average temperature level," which you'll need to create.

Start by computing the states' average temperature. Use that measurement to classify them according to the following criteria:

-   low: temp \< 20
-   Mid: temp \>= 20 and temp \< 25
-   High: temp \>= 25

```{r}
# Average temperature for each state
state_avg_temp <- dat %>%
  group_by(STATE) %>%summarize(
    avg_temp = mean(temp, na.rm = TRUE),
    avg_wind_speed = mean(wind.sp, na.rm = TRUE),
    avg_atm_pressure = mean(atm.press, na.rm =  TRUE))

# Create a new column to store categorical values
state_avg_temp <- state_avg_temp %>%
  mutate(avg_temp_level = ifelse(avg_temp < 20, "Low",ifelse(avg_temp >= 20 & avg_temp < 25, "Mid", "High")))

summary_table <- state_avg_temp %>% group_by(avg_temp_level) %>%summarize(
    num_states = n(),
    avg_temp = mean(avg_temp,na.rm = TRUE),
    avg_wind_speed = mean(avg_wind_speed,na.rm = TRUE),
    avg_atm_pressure = mean(avg_atm_pressure,na.rm = TRUE))

summary_table
state_avg_temp
```

Once you are done with that, you can compute the following:

-   Number of entries (records),
-   Number of NA entries,
-   Number of stations,
-   Number of states included, and
-   Mean temperature, wind-speed, and atmospheric pressure.

All by the levels described before.

```{r}
# I don't know if my output for question 3 is right, but some of the information in Q3 is not included and not used for Q4 (ex. State and USAFID since I grouped them in Q3)
# Define temperature levels
temperature_levels <- c("Low", "Mid", "High")
summary_table1 <- state_avg_temp %>%filter(avg_temp_level %in% temperature_levels) %>%
  group_by(avg_temp) %>% summarize(
    num_entries = n(),
    num_na_entries = sum(is.na(avg_temp)),
    num_states = n_distinct(STATE),
    mean_temp = mean(avg_temp, na.rm = TRUE),
    mean_wind_speed = mean(avg_wind_speed, na.rm = TRUE),
    mean_atm_pressure = mean(avg_atm_pressure, na.rm = TRUE))
summary_table1
```

Knit the document, commit your changes, and push them to GitHub. Once you're done, you can email me the link to your repository or, preferably, link to the [Lab 5 Issue](https://github.com/USCbiostats/PM566/issues/68) in your commit message, as follows:

``` bash
git commit -a -m "Finalizing lab 5 https://github.com/USCbiostats/PM566/issues/68"
```

This will let me know which version of your repository to look at for grading purposes.
